Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-14 MScDiplom_Lemenkova.tex
Comments of Valentijn: 6,10,16,17. Updated image of sensors ASTER, MODIS, MERIS, SeaWiFS (now with spectral bands)

@@

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} (Fig.\ref{fig:1})\label{page-3} are subjected to the human activities\index[concepts]{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. 
One of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06}\label{Leoni07}. 
The detailed research of the fish farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays \textit{P.oceanica} is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution\index[concepts]{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.

@@

Besides anthropogenic factors, various biochemical, climatic and environmental processes\index[concepts]{Seagrasses!natural impacts on} can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature\label{page-4}; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.

@@

\subsection{Research objective}\index[concepts]{Research!objective}
The current MSc research aims to explore the environmental conditions for the spatial distribution of \textit{P.oceanica} seagrass along the northern coast of Crete Island, based on the remote sensing and \ac{GIS} techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources \index[concepts]{Data}:
	\begin{itemize}
		\item[] i) spectra of \textit{P.oceanica}, carbonate sand, silt and other seafloor types
		\item[] ii) satellite imagery: \ac{Landsat TM}, \ac{Landsat ETM+}
		\item[] iii) aerial photos: Google Earth
		\item[] iv) \textit{in-situ} fieldwork data of underwater videographic measurements
		\item[] v) vector GIS layers.
	\end{itemize}
Although there are a variety of environmental factors that contribute to the spectral
reflectance, the most important ones are water column height and seafloor fraction.
It is because spectra of the seagrass \textit{P.oceanica} vary qualitatively over the depths interval of 0.5–4 m,
and secondly, the content and cover fraction of the seafloor have the most distinctive effect on the
spectral reflectance of the water. \\
Therefore, in the current research we focus on these two major factors, and study the response of the
water reflectance towards changing conditions of the water column depth and seafloor bottom cover
fraction (seagrass \textit{P.oceanica}, carbonate sand and silt).\\
The research aims to explore the limitations of the application of the \ac{RTM} and remote sensing techniques 
towards the study of the environmental properties of the \textit{P.oceanica} meadows distribution
over the selected locations of northern Crete. The main research objective is monitoring the seagrass \textit{P.oceanica} in selected areas of Crete and analyzing the environmental conditions for the growth of \textit{P.oceanica}. \\
This research is supported by the \textit{in-situ} measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: spectral modelling by means of the \ac{RTM} \ac{WASI}, underwater videometric measurements made by \ac{Olympus ST} camera, aerial Google Earth and satellite images from different sources, spatial \ac{GIS} and statistical analysis.

@@

The pre-processing of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).
\pagebreak

@@

\subsection[\textit{In-situ} observations of the seagrass]{The \textit{in-situ} observations of the seagrass}
The traditional methods of \textit{in-situ} seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03}\label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, \ac{GPS}, laptops with installed software (\ac{GIS}), \ac{SCUBA} equipment, \ac{iPAQ} and other auxiliary tools\label{page-15} (Fig.\ref{fig:3.3}). The seagrass sampling is taken on the regular way with observation points covering the study area with normal distribution.
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the \textit{in-situ} seagrass observations of the structuring epiphyte
community composition in the \textit{P.oceanica} ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming \textit{in-situ} observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}.  The results of the \textit{in-situ} measurements and observation are usually
managed and treated using integrated \ac{GIS} approach, as, e.g. reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.\\
Other methods of seagrass \textit{in-situ} monitoring are based on the application of the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then collect the return echoes for the analysis (Fig.\ref{fig:sonar}). One of the examples of the acoustic sonar sensor designed for seabed classification is the British \href{http://sonavision.co.uk/products.asp?cat_id=1}{RoxAnn} system, which application for the seafloor mapping is \href{http://www.ozcoasts.org.au/geom_geol/toolkit/Tech_CA_asc.jsp}{very well described} on the \href{http://www.ozcoasts.org.au/}{OzCoasts website}.\\ However, the applications of the acoustic methods requires specialized expensive equipment and is mostly used in the deep open waters, combined with bathymetric measurements. Another limitation of the acoustic
techniques is that, initially intended for the bathymetric surveying, acoustic equipment is mostly designed for the geomorphological and geological studies of the underwater substratum and are, therefore, more adjusted for the benthic habitat discrimination, and they are not effective for the identifying of the biological species composition or even the presence of aquatic vegetation such as seagrasses and seaweeds \cite{Werdell03}\label{Werdell03}.\\
The current study is based on the application of the remote sensing optical measurement techniques, due to their effectiveness, non-destructive nature and availability of necessary tools: spectral radiometers and \ac{RTM}.

@@

Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, \ac{GPS} data, remote sensing and \ac{SCUBA}-based (Fig.\ref{fig:A.52}) fieldwork measurements.

\ac{SCUBA}-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping. 
The underwater videography\index[concepts]{Remote sensing!underwater videography} with a \ac{GPS}\index[concepts]{Remote sensing!GPS} is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}. 
However, it cannot cover large areas for small-scale mapping.
Remote sensing techniques offer clear advantages over other methods of \textit{in-situ} field measurements
and seagrass observations, mentioned above. Preference of the remote sensing methods consists in
their weather-independence, cost-effectiveness, accuracy and spatial coverage, which enables
periodic monitoring of the seagrass meadows and gives access to the distant and unapproachable
areas. 
Integrated together with \ac{GIS} vector layers and maps, remote sensing data enable historical
mapping \cite{Carter08,Ardizzone06}\label{Carter08} \label{Ardizzone06} and assessment of change detection.

@@

However, application of the remote sensing techniques for mapping of submerged vegetation,
seafloor cover types and benthic vegetation, \textit{inter alia} seagrasses, are still in their development.
Approaches and methods for the seagrass protection and monitoring still remain location-specific or, at
least, nation-specific, depending to large extent on the tools available for the researchers \cite{Mellors09a}\label{Mellors09a}.
Universal, international, standardized methods for seagrass directly for seagrasses
as such still should be developed. Various case studies have been performed, yet their mostly report
methods adjusted for particular areas, without evaluating standard general algorithms that could be
extrapolated towards other regions. \\ Application of the remote sensing towards seagrass mapping is
generally based on the assumption that various types of the seafloor bottom have different
characteristics of the reflectivity, which is visually expressed in distinct colors of the objects. In its
turn, reflectivity of the sediments is affected by the water optical properties and content. For
example, Stephens \cite{Stephens03}\label{Stephens03} prove that microalgae biomass and community structure affect
hyperspectral reflectance of sediments, which enable to estimate total microalgae biomass from
measurements of hyperspectral reflectance. 
Spectral measurements\index[concepts]{Remote sensing!spectral measurements} of the target objects are made
by means of the radiometers (Fig.\ref{fig:2.1})\label{page-17}, which receive and register the amounts of energy (radiance and
irradiance) from the objects. Measuring optical properties of of the seawater allows to calculate spectra of the
objects and to discriminate them on the aerial and satellite images. Thus, various scientists report
success in spectral discrimination of submerged vegetation and other seafloor cover types on imagery
using hyperspectral optical properties of the sea water for the assessment of benthic habitats \cite{Lewis01,Louchard03,Dehouck08,Werdell03}\label{Lewis01}\label{Louchard03}\label{Dehouck08}\label{Werdell03}.

@@

Perhaps, the most known imagery, widely used in the remote sensing mapping, is received from the
group of Landsat sensors, known for its historical, pioneer role in the satellite industry.
Being the longest running satellite system, lunched in 1972, Landsat is the only source of archival
data going back to 1984 at a sufficient spatial resolution \cite{Dekker05b}\label{Dekker05b}, which makes its data
desirable for historical mapping or environmental analysis of change detection of the seagrass
landscapes.
The \ac{Landsat TM}\index[concepts]{Remote sensing!multispectral imagery!Landsat TM} and \ac{Landsat ETM+}\index[concepts]{Remote sensing!multispectral imagery!Landsat ETM+} data, with the most recent from sensor Landsat 7 - an advanced
and multispectral scanning, launched in 1999, prove to be feasible and useful for the mapping of
submerged vegetation, such as seagrasses or coral reefs. The successful applications of the imagery
\ac{Landsat TM} towards the seagrass mapping, were reported in numerous research works \cite{Palandro03,Gullstroom06,Wabnitz08,Bierwirth93,Ferguson97,Rasib97}\label{Palandro03}\label{Gullstroom06}\label{Wabnitz08}\label{Bierwirth93}\label{Ferguson97}\label{Rasib97}.\\ The Landsat data are particularly suitable for the
case of change detection of seagrass landscapes at a decade scale, because being the main sensor
onboard the Landsat satellites, the Thematic Mapper (TM) provides the longest time series
available for change detection analysis over submerged vegetation \cite{Palandro03}\label{Palandro03}.

@@

Another well-known multispectral sensor, \ac{SPOT}\index[concepts]{Remote sensing!multispectral imagery!SPOT} provides multispectral imagery with a spatial resolution of 10 m (Fig.\ref{fig:2.4}), covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \ac{SPOT} imagery was used for mapping beds of \textit{P.oceanica} in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}. 
The \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}\index[concepts]{Remote sensing!multispectral imagery!IKONOS}, offering multispectral and panchromatic imagery, was the first to collect publicly
available high-resolution imagery at 1- and 4-meter resolution from \href{http://www.geoeye.com}{Geoeye}. IKONOS
imagery has been applied for the seagrass mapping due to its high resolution and accessibility. 
Thus, the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, \ac{Landsat TM}, and \ac{CASI}\index[concepts]{Remote sensing!multispectral imagery!CASI}, show that in the blue part of the
spectrum, the best results are achieved by the IKONOS and \ac{CASI}, while \ac{Landsat TM} has not high
enough resolution. 
It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that \ac{CASI} enable to receive still more accurate results of the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.

@@

The information about the location of the seagrass (mostly represented by the 	extit{P.oceanica} species) is useful for the

@@

\begin{figure}
	\centering
	\subfloat [small-scaled (1:54,000)]{\label{fig:38a}\includegraphics[width=0.8\textwidth]{Fig35a.jpg}}\\
	\subfloat [enlarged (1:10,000)]{\label{fig:38b}\includegraphics[width=0.8\textwidth]{Fig35b.jpg}}
	\caption{Seafloor types of the Arina beach: \textit{P.oceanica}, sand, rocks. \\ \href{http://www.google.com/earth/index.html}{Google Earth} images. Page: $<$\pageref{page-48}$>$}
	\label{fig:38}
\end{figure}

>>>>

	\acro{GDAL}{\href{http://www.gdal.org/}{Geospatial Data Abstraction Library}}

@@

	\acro{HCMR}{\href{http://www.hcmr.gr/}{Hellenic Centre for Marine Research }}

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} (Fig.\ref{fig:1})\label{page-3} are subjected to the human activities\index[concepts]{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. 
One of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading \cite{Cambridge09}\label{Cambridge09}, \cite{Leoni08}\label{Leoni08}, which may occasionally casue morphological (e.g. leaf length) and physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06}\label{Leoni07}. 
The detailed research of the fish farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays \textit{P.oceanica} is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution\index[concepts]{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.

@@

Besides anthropogenic factors, various biochemical, climatic and environmental processes\index[concepts]{Seagrasses!natural impacts on} can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature\label{page-4}; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses are 
increased nutrient availability in the coastal zones \cite{Alcoverro1997}\label{Alcoverro1997}, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.

@@

\subsection{Research objective}\index[concepts]{Research!objective}
The current MSc research aims to explore the environmental conditions for the spatial distribution of \textit{P.oceanica} seagrass along the northern coast of Crete Island, based on the remote sensing and \ac{GIS} techniques, knowledge about the coastal environment in Crete and integration of various data
from the following sources \index[concepts]{Data}:
	\begin{itemize}
		\item[] i) spectra of \textit{P.oceanica}, carbonate sand, silt and other seafloor types
		\item[] ii) satellite imagery: \ac{Landsat TM}, \ac{Landsat ETM+}
		\item[] iii) aerial photos: Google Earth
		\item[] iv) \textit{in-situ} fieldwork data of underwater videographic measurements
		\item[] v) vector GIS layers.
	\end{itemize}
Although there are a variety of environmental factors that contribute to the spectral
reflectance, the most important ones are water column height and seafloor fraction.
It is because spectra of the seagrass \textit{P.oceanica} vary qualitatively over the depths interval of 0.5–4 m,
and secondly, the content and cover fraction of the seafloor have the most distinctive effect on the
spectral reflectance of the water. 
In the environmental conditions of Crete Island, \textit{P.oceanica} may grow at depths up to six meters; however, the most usual depth is four meters, 
which caused our decision to focus on depths up to four meters, and use these depths for further WASI modelling. 
Therefore, in the current research we focus on these two major factors, and study the response of the
water reflectance towards changing conditions of the water column depth and seafloor bottom cover
fraction (seagrass \textit{P.oceanica} and carbonate sand).\\
The research aims to explore the limitations of the application of the \ac{RTM} and remote sensing techniques 
towards the study of the environmental properties of the \textit{P.oceanica} meadows distribution
over the selected locations of northern Crete. The main research objective is monitoring the seagrass \textit{P.oceanica} in selected areas of Crete and analyzing the environmental conditions for the growth of \textit{P.oceanica}. \\
This research is supported by the \textit{in-situ} measurements in two selected locations of the northern coast
of Crete (Ligaria beach in Agia Pelagia district and Xerocampos), using following methods of the
remote sensing techniques: spectral modelling by means of the \ac{RTM} \ac{WASI}, underwater videometric measurements made by \ac{Olympus ST} camera, aerial Google Earth and satellite images from different sources, spatial \ac{GIS} and statistical analysis.

@@

The pre-processing of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution (Fig.\ref{fig:A.50})\label{GG-2}, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).
\pagebreak

@@

\subsection[\textit{In-situ} observations of the seagrass]{The \textit{in-situ} observations of the seagrass}
The traditional methods of \textit{in-situ} seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03}\label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, \ac{GPS}, laptops with installed software (\ac{GIS}), \ac{SCUBA} equipment, \ac{iPAQ} and other auxiliary tools\label{page-15} (Fig.\ref{fig:3.3}). The seagrass sampling is taken on the regular way with observation points covering the study area with normal distribution.
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the \textit{in-situ} seagrass observations of the structuring epiphyte
community composition in the \textit{P.oceanica} ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming \textit{in-situ} observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}.  The results of the \textit{in-situ} measurements and observation are usually
managed and treated using integrated \ac{GIS} approach, as, e.g. reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.\\
Other methods of seagrass \textit{in-situ} monitoring are based on the application of the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then collect the return echoes for the analysis (Fig.\ref{fig:sonar}).

One of the examples of the acoustic sonar sensor designed for seabed classification is the British \href{http://sonavision.co.uk/products.asp?cat_id=1}{RoxAnn} system, which application for the seafloor mapping is \href{http://www.ozcoasts.org.au/geom_geol/toolkit/Tech_CA_asc.jsp}{very well described} on the \href{http://www.ozcoasts.org.au/}{OzCoasts website}.\\ However, the applications of the acoustic methods requires specialized expensive equipment and is mostly used in the deep open waters, combined with bathymetric measurements. Another limitation of the acoustic
techniques is that, initially intended for the bathymetric surveying, acoustic equipment is mostly designed for the geomorphological and geological studies of the underwater substratum and are, therefore, more adjusted for the benthic habitat discrimination, and they are not effective for the identifying of the biological species composition or even the presence of aquatic vegetation such as seagrasses and seaweeds \cite{Werdell03}\label{Werdell03}.\\
The current study is based on the application of the remote sensing optical measurement techniques, due to their effectiveness, non-destructive nature and availability of necessary tools: spectral radiometers and \ac{RTM}.

@@

Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, \ac{GPS} data, remote sensing and \ac{SCUBA}-based (Fig.\ref{fig:A.52}) fieldwork measurements.
\ac{SCUBA}-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping. 
The underwater videography\index[concepts]{Remote sensing!underwater videography} with a \ac{GPS}\index[concepts]{Remote sensing!GPS} is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}. 
However, it cannot cover large areas for small-scale mapping.
Remote sensing techniques offer clear advantages over other methods of \textit{in-situ} field measurements
and seagrass observations, mentioned above. Preference of the remote sensing methods consists in
their weather-independence, cost-effectiveness, accuracy and spatial coverage, which enables
periodic monitoring of the seagrass meadows and gives access to the distant and unapproachable
areas. 
Integrated together with \ac{GIS} vector layers and maps, remote sensing data enable historical
mapping \cite{Carter08,Ardizzone06}\label{Carter08} \label{Ardizzone06} and assessment of change detection.

@@

However, application of the remote sensing techniques for mapping of submerged vegetation,
seafloor cover types and benthic vegetation, \textit{inter alia} seagrasses, are still in their development.
Approaches and methods for the seagrass protection and monitoring still remain location-specific or, at
least, nation-specific, depending to large extent on the tools available for the researchers \cite{Bakran-Petricioli06}\label{Bakran-Petricioli06}, \cite{Mellors09a}\label{Mellors09a}.
Universal, international, standardized methods for seagrass directly for seagrasses
as such still should be developed. Various case studies have been performed, yet their mostly report
methods adjusted for particular areas, without evaluating standard general algorithms that could be
extrapolated towards other regions. \\ Application of the remote sensing towards seagrass mapping is
generally based on the assumption that various types of the seafloor bottom have different
characteristics of the reflectivity, which is visually expressed in distinct colors of the objects \cite{Richardson06}\label{Richardson06}. In its
turn, reflectivity of the sediments is affected by the water optical properties and content. For
example, Stephens \cite{Stephens03}\label{Stephens03} prove that microalgae biomass and community structure affect
hyperspectral reflectance of sediments, which enable to estimate total microalgae biomass from
measurements of hyperspectral reflectance. 
Spectral measurements\index[concepts]{Remote sensing!spectral measurements} of the target objects are made
by means of the radiometers (Fig.\ref{fig:2.1})\label{page-17}, which receive and register the amounts of energy (radiance and
irradiance) from the objects. Measuring optical properties of of the seawater allows to calculate spectra of the
objects and to discriminate them on the aerial and satellite images. Thus, various scientists report
success in spectral discrimination of submerged vegetation and other seafloor cover types on imagery
using hyperspectral optical properties of the sea water for the assessment of benthic habitats \cite{Lewis01,Louchard03,Dehouck08,Werdell03}\label{Lewis01}\label{Louchard03}\label{Dehouck08}\label{Werdell03}.

@@

Perhaps, the most known imagery, widely used in the remote sensing mapping, is received from the
group of Landsat sensors, known for its historical, pioneer role in the satellite industry.
Being the longest running satellite system, lunched in 1972, Landsat is the only source of archival
data going back to 1984 at a sufficient spatial resolution \cite{Dekker05b}\label{Dekker05b}, which makes its data
desirable for historical mapping or environmental analysis of change detection of the seagrass
landscapes.
The \ac{Landsat TM}\index[concepts]{Remote sensing!multispectral imagery!Landsat TM} and \ac{Landsat ETM+}\index[concepts]{Remote sensing!multispectral imagery!Landsat ETM+} data, with the most recent from sensor Landsat 7 - an advanced
and multispectral scanning, launched in 1999, prove to be feasible and useful for the mapping of
submerged vegetation, such as seagrasses or coral reefs. The successful applications of the imagery
\ac{Landsat TM} towards the seagrass mapping, were reported in numerous research works \cite{Palandro03,Gullstroom06,Wabnitz08,Bierwirth93,Ferguson97,Rasib97}\label{Palandro03}\label{Gullstroom06}\label{Wabnitz08}\label{Bierwirth93}\label{Ferguson97}\label{Rasib97}.\\ The Landsat data are particularly suitable for the
case of change detection of seagrass landscapes at a decade scale, because being the main sensor
onboard the Landsat satellites, the Thematic Mapper (TM) provides the longest time series
available for change detection analysis over submerged vegetation \cite{Palandro03}\label{Palandro03}.
Another well-known multispectral sensor, \ac{SPOT}\index[concepts]{Remote sensing!multispectral imagery!SPOT} provides multispectral imagery with a spatial resolution of 10 m (Fig.\ref{fig:2.4}), covering covers a surface area of 3600 km² (60*60 km swath), 26-day
orbital repeat cycle for nadir viewing and imagery with a spatial resolution 20 - 2.5 m
\cite{SPOT}\label{SPOT}. \ac{SPOT} imagery was used for mapping beds of \textit{P.oceanica} in the
Mediterranean Sea \cite{Pasqualini05} \label{Pasqualini05}. 
The \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}\index[concepts]{Remote sensing!multispectral imagery!IKONOS}, offering multispectral and panchromatic imagery, was the first to collect publicly
available high-resolution imagery at 1- and 4-meter resolution from \href{http://www.geoeye.com}{Geoeye}. IKONOS
imagery has been applied for the seagrass mapping due to its high resolution and accessibility. 
Thus, the results of image classification in case study of shallow-water marine environment \cite{Mumby02}\label{Mumby02}
made using IKONOS, \ac{Landsat TM}, and \ac{CASI}\index[concepts]{Remote sensing!multispectral imagery!CASI}, show that in the blue part of the
spectrum, the best results are achieved by the IKONOS and \ac{CASI}, while \ac{Landsat TM} has not high
enough resolution. 
It may be caused to some extent by the loss of the radiance contrast, atmospheric
Rayleigh scattering and defects of scattering. However, comparing between CASI and IKONOS, the
same authors prove that \ac{CASI} enable to receive still more accurate results of the classification
than IKONOS \cite{Mumby02}\label{Mumby02}.

@@

Another comparative analysis of the application of \href{http://www.itres.com/products/imagers/casi550}{CASI}, Landsat and \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird} imagery \cite{Phinn08}\label{Phinn08}
 demonstrates high suitability of \ac{CASI} images for the fine-scale mapping of the seagrass
landscapes. Thus, \ac{CASI} and Quickbird-2\index[concepts]{Remote sensing!multispectral imagery!Quickbird} images enable to identify even separate seagrass species
with small width and heterogeneous nature of the seagrass patches, which could not be detected using
Landsat TM images with their 30*30m resolution.
\ac{ASTER}\index[concepts]{Remote sensing!multispectral imagery!ASTER}), launched in
1999 on board Terra sensor\index[concepts]{Remote sensing!multispectral imagery!Terra sensor}, provides high-resolution images of the Earth in 15 different bands of the
electromagnetic spectrum, ranging from visible to thermal infrared light  (Fig.\ref{fig:2.4}). It has diverse
subsystems for the \ac{VNIR} with 15 m resolution, shortwave infrared \ac{SWIR}, and
\ac{TIR} wavelength regions \cite{Kalinowski04}\label{Kalinowski04}.

@@

The information about the location of the seagrass (mostly represented by the \textit{P.oceanica} species) is useful for the

@@

\begin{figure}[H]
	\centering
	\subfloat [small-scaled (1:54,000)]{\label{fig:38a}\includegraphics[width=0.45\textwidth]{Fig35a.jpg}}
	\hspace{1ex}
	\subfloat [enlarged (1:10,000)]{\label{fig:38b}\includegraphics[width=0.45\textwidth]{Fig35b.jpg}}
	\caption{Seafloor types of the Arina beach: \textit{P.oceanica}, sand, rocks. \\ \href{http://www.google.com/earth/index.html}{Google Earth} images. Page: $<$\pageref{page-48}$>$}
	\label{fig:38}
\end{figure}

On the basis of the image processing and classification, applied to the northern coats of Crete, the
marine bottom types between 10 m depth has been mapped, including the limit depths of the of the
seagrass meadows provided on the basis of the available and collected field data. The classification
has been completed using Erdas Imagine software.
\pagebreak
